---
title: "From space to coast: a semi-automatic quantative review on remote sensing of coastal ecosystems"
author: "RS-eco"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(comment = NA, warning=F, message=F, echo=F, 
                      fig.path="figures/", dpi = 1000)
```

<!-- Check out: Johnson et al. 2013, Andradi-Brown et al. 2013 & Bernd et al. 2016 -->

## Abstract

<!-- Need to re-write abstract, identical to introduction -->
<!-- Shorten intro, lengthen methods & results, add discussion! -->
Direct sampling of marine communities can be very difficult. Thus, marine research intensively relies on indirect observation methods, such as tethered video surveys or the use of Remotely Operated Vehicles (ROVs) and Autonomous Underwater Vehicles (AUVs). Bathymetry and environmental data are often gathered through ship-based sensors or airborne and satellite systems, which allow for continuous sampling over large spatial and temporal scales. In situ observations using unmanned aerial vehicles (UAVs), camera traps and acoustic sensors provide another indirect sampling method. All these methods have been increasingly used and have helped to acquire a vast amount of data, which helps us to further our scientific understanding and make predictions about future changes in marine ecosystems. Although, coastal communities are more accessible than most of the marine realm and can thus also be easily studied without remote sensing, the accessibility also means that they are more easily threatened by anthropogenic activities, such as coastal pollution and land use change. Thus, coastal ecosystems are an important research area, which are frequently studied using direct sampling but also remote sensing. Here, a quantitative literature review of the current literature on remote sensing and coastal ecosystems was performed. A total number of 3835 peer-reviewed journal publications was identified using Web of Knowledge and significant patterns and trends in these publications were identified. There was an exponential increase in the number of publications from 1990 - 2017, with a particular sharp increase since 2000. Most studies on remote sensing in coastal ecosystems have been conducted in China, India, Mali, Australia, Brazil, North America and Mexico. There was a strong difference in the number of publications by research topic, as well as by coastal ecosystem and taxonomic group. Most studies further utilised multispectral imagery, rather than hyperspectral, lidar or radar data and also mostly used Modis and Landsat imagery. ...

**Keywords:** marine, satellite, biodiversity, conservation, climate change, pollution

```{r pubs}
# Load packages
library(dplyr); library(readr); library(ggplot2)

# List files
fl <- list.files("data/abstracts", pattern=".txt", full.names = T)

# Identify search terms
search_terms <- unique(sapply(fl, function(x) paste(strsplit(basename(x), split="_")[[1]][1:2], collapse="_")))

# Read files by search term
d <- lapply(search_terms, function(x){
  files <- list.files("data/abstracts", pattern=x, full.names = T)
  d <- lapply(files, read_tsv)
  d <- do.call("rbind", d)
  d <- select(d, c(AU, TI, SO, PY, AB))
  d$search <- x
  return(d)
})

# Merge files
d <- Reduce(function(x,y) dplyr::full_join(x,y, by=c("AU", "TI", "SO", "PY", "AB")), d)

# Join search columns to one columns
d <- tidyr::unite(d, "search", contains("search"), sep=" ")
d$search <- sapply(d$search, function(x){
   paste0(unique(strsplit(x, split=" ")[[1]])[unique(strsplit(x, split=" ")[[1]]) != "NA"], collapse=", ")
})

# Select number of unique studies
d <- unique(d)
d_all <- d

# Could there be any duplicates if year is wrong/abstract missing, ...

# Plot total number of publications per year
#d %>% group_by(PY) %>% summarise(n = n()) %>%
#ggplot(aes(x=PY, y=n)) + geom_point() + geom_line() + 
#  scale_x_continuous(limits=c(1970, 2017)) + theme_bw() + 
#  labs(x="Year", y="Number of Publications")

#Extent keywords for identifying marine and terrestrial studies, needed?

# Change ALL CAPITAL TEXT to Normal
d$TI <- sapply(d$TI, tolower)
d$AB <- sapply(d$AB, tolower)

# Identify terrestrial, marine and coastal studies
d$marine[grepl(x=d$TI, pattern="terrestrial") | grepl(x=d$AB, pattern="terrestrial") | 
           grepl(x=d$TI, pattern="land") | grepl(x=d$AB, pattern="land")] <- "terrestrial"
d$marine[grepl(x=d$TI, pattern="marine") | grepl(x=d$AB, pattern="marine") | 
         grepl(x=d$TI, pattern="deep-sea") | grepl(x=d$AB, pattern="deep-sea")] <- "marine"
d$marine[grepl(x=d$TI, pattern="coast") | grepl(x=d$AB, pattern="coast") | 
           grepl(x=d$TI, pattern="shore") | grepl(x=d$AB, pattern="shore") | 
          grepl(x=d$TI, pattern="estuar") | grepl(x=d$AB, pattern="estuar") | 
           grepl(x=d$TI, pattern="intertidal") | grepl(x=d$AB, pattern="intertidal")] <- "coastal"

# Plot number of publications over time and by marine
#d %>% select(-TI, -AU, -SO) %>% 
#  group_by(PY, marine) %>% 
#  summarise(sum = n()) %>% tidyr::drop_na() %>% ungroup() %>%
#  ggplot(aes(x=PY, y=sum, colour=marine)) + geom_point() + geom_line() + 
#  theme_bw() + labs(x="Year", y="Number of Publications") + 
#  scale_x_continuous(limits=c(1990, 2017), expand=c(0,0)) + 
#  scale_y_continuous(limits=c(0,500), expand=c(0,0)) + scale_colour_discrete(name="") + 
#  theme(legend.position=c(0.1,0.87))

# Only select coastal studies
d <- d %>% filter(marine == "coastal")
```

## Introduction

Global change is a hot topic in environmental and ecological research. It imposes challenges not only to the research, but also to the management and policy community and these challenges are likely to increase even further in the future (Pereira et al. 2010). Understanding current and making reliable predictions of future changes, enables us to prepare for these challenges, but also requires the collection of a vast amount of data.  

Collecting field data is generally very resource intensive and difficult to implement across large spatial and temporal scales. Remote Sensing (RS) offers a good alternative and thus is an important tool for measuring environmental conditions, as well as the state of biological diversity and ecosystem services across multiple spatial and temporal scales. Thus, it is not surprising that the use of RS has experienced an exponential increase over the past decade (Pettorelli et al. 2014). Sensors, methodologies and data availability have strongly developed and using RS has become simpler and more widespread (Palmer et al. 2014). This is not only true for the terrestrial environment, but RS has also become increasingly important in studying coastal and marine ecosystems (Brewington et al. 2014).   
The direct sampling of coastal and marine communities is much harder than the direct sampling of terrestrial communities. Thus, coastal and marine research intensively relies on indirect observation methods, such as tethered video surveys (Biber et al. 2014) or the use of Remotely Operated Vehicle (ROVs; Lorance & Trenkel 2006) and Autonomous Underwater Vehicle (AUVs; Eriksen et al. 2001, Yoerger et al. 2007). Bathymetry and environmental data are often gathered through ship-based sensors or airborne (Scheritz & Dietrich 2002, Koh & Wich 2012) and satellite systems (Baldock et al. 2014), which allow for continuous sampling over large spatial and temporal scales. In situ observations using unmanned aerial vehicles (UAVs)<!--Ref-->, camera traps (Bailey et al. 2007, Ahumada et al. 2011) and acoustic sensors (Van Parijs et al. 2009) provide another indirect sampling method.  

<!-- Field spectrometry (Murphy et al. 2008), as well as vessel monitoring systems (Bastardie et al. 2010) -->

These methods have already been applied in a large range of areas of coastal and marine research (Brewington et al. 2014). Environmental variables, such as sea surface temperature (Smit et al. 2013, Baldock et al. 2014), chlorophyll-α (Dierssen 2010, Raitsos et al. 2013), salinity (Lagerloef et al. 1995, 2008) and suspended particulate matter concentration (Evans et al. 2012, Bowers et al. 2014) can be easily obtained from satellite measurements and are frequently used in marine environmental research. Sea floor features and bathymetry have been widely mapped using ship-based depth soundings and satellite altimetry (Smith 1997, Becker et al. 2009). Remote sensing can also be used for shoreline change detection (Chen & Rau 1998, Li et al. 2001), mapping of coastal habitats, such as salt marshes (Ozesmi & Bauer 2002, Belluco et al. 2006), coral reefs (Lirman et al. 2007, Huvenne et al. 2011), as well as seagrass (Pasqualini et al. 2001) and mangrove ecosystems (Giri et al. 2011, Satyanarayana et al. 2011), fisheries management (Stuart et al. 2011), species movements (Lander et al. 2013) and risk assessment (Gillespie et al. 2007, Römer et al. 2012).  

Although, coastal communities are more accessible than most of the marine realm and so can be studied much easier than for example the deep-sea, the better accessibility also means that they are more easily threatened by anthropogenic activities. A significant proportion of the global human population lives in close proximity to the coast and so causes the pollution and land degradation of coastal ecosystems<!--Ref-->. Thus, coastal ecosystems are an important research area, which are frequently studied using direct sampling but also remote sensing.  

<!--
Coastal ecosystems, such as mangroves, salt marshes, corals and seagrasses, provide substantial benefits to humanity, yet are under ever-increasing threat of degradation and disappearance from a combination of land use and climate change (such as sea level rise, drought and storm intensity). With about 10% of the global human population living in the vulnerable low-elevation coastal zone, mitigation and management of threats to coastal ecosystems is high on the climate change resilience agenda. Over the past decades, remote sensing techniques have enabled monitoring of coastal ecosystem functioning, threats and opportunities for improved management of these often-inaccessible systems.
-->

Rose et al. (2014) identified 10 topics, where using remotely sensed data would help to answer important conservation questions. These topics are: species distributions and abundances, species movements and life stages, ecosystem processes, climate effects, rapid response, protected areas, ecosystem services, conservation effectiveness, agricultural and aquacultural expansions and changes in land use land cover (LULC) and degradation and disturbance regimes. These topics highlight the importance of RS not only for conservation but also for environmental research in general, especially with regard to climate and land-use change.  

In the following, trends and patterns in the current literature on remote sensing and coastal ecosystems will be analysed. I will further highlight, which of the 10 topics, identified by Rose et al. (2014), are already intensively studied in coastal ecosystems and where further research is required in the future in order to provide essential information for mitigating future effects of global change.  

<!-- Importance of coastal ecosystems! Talk about threats, coastal pollution,land degradation, ... -->

## Materials and Methods

A literature search for peer-reviewed publications, published in English over the period 1990–2018, describing coastal studies using remote sensing was performed. The literature search was undertaken using the commercial search engine Web of Science. Combinations of the following search terms were used: ‘remote sensing’, ‘coast’, ‘marine’, ‘landsat’, ‘MODIS’, ‘Sentinel’, ‘seagrass’, ‘corals’, ‘mangroves’, ‘salt marsh’, ‘salt marshes’, ‘biodiversity’ and ‘conservation’. For a list of the exact search terms and the resulting number of publications see Table S1. The number of publications that overlapped among the different search terms is further depicted in Fig. S1. Only peer-reviewed articles were chosen as they form the main body of literature widely available to researchers. The articles used therefore did not include any university theses, technical governmental reports or conference proceedings.
<!-- Really overlapped? Rather covered? -->
<!-- 
Extend search to: remote sensing, remote sensing* terrestrial, landsat, modis, 
Sentinel, Grace, Spot, RapidEye, Aquarius, Pleiades, SeaWifs
-->

All resulting database records were downloaded, including the abstract of each publication (where available), and stored in a CSV file. The data was then analysed using the statistical software R (R Core Team 2020). The title and abstract of each study was scanned for various keywords to extract various categorical data variables. First, all studies were grouped into coastal, marine and terrestrial studies. Only coastal studies were used for subsequent analysis. All coastal titles and abstracts were scanned for country names to identify in which country each study was located. Furthermore, the value of six categorical variables were extracted from every coastal study. The variables were study theme (pollution, conservation, climate change, biodiversity), research topic (10 topics identified by Rose et al. (2014)), coastal ecosystem (coral reefs, estuaries, intertidal zones, kelp forests, mangroves, mudflats, rocky shores, salt marshes, sandy shores, seagrass meadows, wetlands), taxonomic group (birds, fishes, invertebrates, mammals, reptiles), sensor (multispectral, hyperspectral, radar, lidar) and EO satelllite (Aquarius, Aster, Envisat, GOCI, Grace, Ikonos, Landsat, MODIS, Pleiades, RapidEye, SeaWifs, Sentinel, Spot). The list of keywords used to identify the corresponding value of each category can be found in Table S2.

<!--Study theme was further divided into sub-categories:  -->
<!-- Problem with country of publisher?-->

<!--The total spatial scale (total area) of each study was calculated using ARC GIS (version 9.2) if it was not described clearly by the authors within the methods.-->

<!-- For each category articles that did not result in any value we individually analysed the abstracts to complete the data entries.-->

<!-- 
Add additional sensors: JPSS-2, TSIS-2, RBI, OMPS-Limb, NI-SAR, PACE, SWOT, TEMPO, GRACE-FO, ICESat-2, CYGNSS, SAGE III, SMAP, OCO-2, TRMM, QuikSCAT, EO-1, Landsat-7/Landsat 7, ACRIMSAT, SORCE, Terra, Aqua, Cloudsat, CALIPSO, Aura, GRACE, OSTM/Jason 2, Aquarius, Suomi NPP, Landsat-8/Landsat 8, GPM, LIS, TSIS-1, GPM, DSCOVR
--->

## Results

The search terms resulted in a total of `r nrow(d_all)` studies of which `r nrow(d)` publications (`r round(nrow(d)/nrow(d_all)*100,0)`%) were related to coastal studies and remote sensing. Of the `r nrow(d)` publications that were analysed, `r nrow(d[is.na(d$AB),])` publications did not contain an abstract.

There was an exponential increase in the number of publications from 1990 - 2017, with a particular sharp increase since 2000. Only from 1999 to 2000 and from 2008 to 2010 there was a decrease in the number of publications. `r round(nrow(d[d$PY >= 2000,])/nrow(d)*100, 0)`% of publications have been published since 2000, and `r round(nrow(d[d$PY >= 2013,])/nrow(d)*100, 0)`% within the last 5 years (Fig. 1, Fig. S2). The majority of studies have been published in four journals (Remote Sensing of Environment, Remote Sensing, Journal of Coastal Research, International Journal of Remote Sensing, Fig. S3).
  
```{r pub, fig.width=8, fig.height=5}
# Identify papers related to biodiversity and conservation
d$`Biodiversity`[grepl(x=d$AB, pattern="biodiversity") | 
                  grepl(x=d$TI, pattern="biodiversity")] <- 1
d$`Conservation`[grepl(x=d$AB, pattern="conservation") | 
                 grepl(x=d$TI, pattern="conservation")] <- 1
d$`Climate Change`[grepl(x=d$AB, pattern="climate change") | 
                 grepl(x=d$TI, pattern="climate change")] <- 1
d$`Pollution`[grepl(x=d$AB, pattern="pollution") | 
                 grepl(x=d$TI, pattern="pollution")] <- 1
d$Sum <- rowSums(d[,c("Biodiversity", "Conservation", "Climate Change", "Pollution")], na.rm=T)
d$None[d$Sum!=0] <- NA
d$None[d$Sum==0] <- 1
d$Multiple[d$Sum>1] <- 1

# Plot number of publications over time
d %>% select(-TI, -AU, -SO) %>% filter(PY < 2017) %>%
  select(PY, Biodiversity, Conservation, `Climate Change`, Pollution, None, Multiple) %>% 
  tidyr::gather(Theme, value, -c(PY)) %>% group_by(PY, Theme) %>% tidyr::drop_na() %>% 
  summarise(sum = n()) %>% ungroup() %>% 
  tidyr::complete(PY, Theme, fill = list(sum=0)) %>% arrange(Theme, PY) %>%
  mutate(Theme = factor(Theme, levels=c("None", "Biodiversity", "Climate Change", "Conservation", "Pollution", "Multiple"))) %>%
  ggplot(aes(x=PY, y=sum, fill=Theme)) + geom_area() + 
  theme_bw() + labs(x="Year", y="Number of Publications") + 
  scale_x_continuous(breaks=seq(1990,2015, by=5), 
                     limits=c(1989, 2016), expand=c(0,0)) + 
  scale_y_continuous(limits=c(0,470), expand=c(0,0)) + 
  scale_fill_manual(values=c('darkgrey','#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e'))
```

**Fig. 1.** Number of coastal publications over time, which are related to remote sensing, until 2016. Number of publications is subdivided by the different research themes (biodiversity, climate change, conservation and pollution). Publications where non of the 4 research themes could be identified are highlighted in grey. See Fig. S2 for number of publications until now.  

```{r country_code}
## Find country of studies

# Load country list
data(codelist, package="countrycode")

# Identify global studies
d$global[grepl(x=d$AB, pattern="global") | grepl(x=d$TI, pattern="global")] <- 1
# Currently also considers studies that mention global change, ...

#Extract country for each study
d$country <- sapply(1:nrow(d), function(x){
  country <- sapply(codelist$country.name.en, function(y){
      if(grepl(d$TI[x], pattern=tolower(y)) | grepl(d$AB[x], pattern=tolower(y))){y}
  })
  country <- Filter(Negate(is.null), country)
  if(length(country)){glue::collapse(country, sep=", ")} else{NA}
})

# Identify number of studies with multiple countries
number_countries <- sapply(d$country, function(x) length(strsplit(x, split=", ")[[1]]))
study_countries <- unlist(lapply(d$country, function(x) strsplit(x, split=", ")[[1]]))

#Create map

#Load countries data for plotting
data(countriesHigh, package="rworldxtra")
library(sf)
countriesHigh <- st_as_sf(countriesHigh)

#Create counts of each country
country_counts <- data.frame(table(study_countries))
country_counts <- na.omit(country_counts)
n_count <- unique(country_counts$study_countries)
n_count_most <- unique(country_counts$study_countries[country_counts$Freq > 250])
n_count_2nd <- unique(country_counts$study_countries[country_counts$Freq >= 100 & country_counts$Freq <= 250])

country_counts <- dplyr::left_join(country_counts, codelist[,c("country.name.en", "iso3c")], by=c("study_countries"="country.name.en"))

# Turn data into categories
country_counts <- na.omit(country_counts)
country_counts$Freq <- cut(country_counts$Freq, c(0, 5, 10, 20, 50, 100, 250, max(country_counts$Freq)))
colnames(country_counts) <- c("Country", "Freq", "ISO_A3")

#Combine country data with map
countriesHigh <- left_join(countriesHigh, country_counts)
```
  
Studies were conducted in `r length(n_count)` different countries. In the majority of countries (85) only a few studies (0-5) have been conducted, while most studies have been conducted in `r unique(n_count_most)` (> 250 studies per country) and `r unique(n_count_2nd)` (100 - 250 studies per country) (Fig. 2).  

<!-- Mali is not even at the coast! Must be a mistake, maybe subname of countries or other words? -->

```{r map, fig.width=8, fig.height=6}
library(patchwork)
countriesHigh <- st_buffer(countriesHigh, dist=0)
# Chloropleth map
a <- ggplot() + geom_sf(data=countriesHigh, aes(fill=Freq)) + 
  #geom_sf_label(data=countriesHigh, aes(label=Freq)) + 
  # Could try to add text with ggrepel
  scale_fill_manual(name="Frequency", na.value="#999999",
                      values=c("#E69F00", "#56B4E9", "#009E73", "#F0E442",
                               "#D55E00", "#0072B2", "#CC79A7")) + 
  cowplot::theme_map() + coord_sf(expand=F, ndiscr=0) + ggtitle("a)")
b <- cowplot::ggdraw(cowplot::get_legend(a))
c <- data.frame(table(countriesHigh$Freq, useNA="ifany")) %>% 
  ggplot(aes(x=Var1, y=Freq)) +
  geom_bar(stat="identity", fill=c("#E69F00", "#56B4E9", "#009E73", "#F0E442", 
                                   "#0072B2", "#D55E00", "#CC79A7", "#999999")) + theme_bw() + 
  geom_text(aes(label=Freq), hjust=-0.5) + ggtitle("b)") + 
  labs(x="Frequency", y="Number of countries") + coord_flip() + 
  scale_y_continuous(limits=c(0,92), expand=c(0,0)) + 
  theme_bw()

{{(a + theme(legend.position="none") | b)} + plot_layout(widths=c(4,1))} / c + plot_layout(heights=c(3,2))
```

**Fig. 2.** (a) Number of studies per country and (b) number of countries with a certain number of studies. `r length(number_countries[number_countries > 1])` studies were conducted in multiple countries. <!--`r sum(d$global, na.rm=T)` studies were conducted globally. !!! Verify number of global studies (global change, ...)!-->
  
<!-- Check global studies, better keywords, add to Table S-->
<!-- Also verify keyword success and add groups that do not have any keyword. -->

Coastal publications covered all major research themes. Most studies focused on conservation (288 publications), climate change and pollution were similarly frequently studied themes (211 and 212 publications), while biodiversity was the least studied theme in coastal remote sensing research (149, Fig. 1 & Fig. S3).   
<!-- Add map of number of studies per country by research theme as subset or add pie chart, ... -->
<!-- Write something on wordcloud results -->

```{r wordcloud, fig.width=8, fig.height=5}
library(tm);library(wordcloud); library(SnowballC)
wordle.df <- rbind(data.frame(doc_id = 1:length(d$AB), text = as.character(d$AB)),
                   data.frame(doc_id = 1:length(d$TI), text = as.character(d$TI)))
wordle.corpus <- Corpus(DataframeSource(wordle.df))
wordle.corpus <- tm_map(wordle.corpus, removePunctuation)
wordle.corpus <- tm_map(wordle.corpus, removeWords, stopwords('english'))
wordle.corpus <- tm_map(wordle.corpus, stripWhitespace)
wordle.corpus <- tm_map(wordle.corpus, tolower)
wordle.corpus <- Corpus(VectorSource(wordle.corpus))
tdm <- TermDocumentMatrix(wordle.corpus, control = list(removePunctuation=TRUE, stopwords = TRUE))
m <- as.matrix(tdm)
v <- sort(rowSums(m),decreasing=TRUE)
words <- data.frame(word = names(v),freq=v)

# Merge title_words and abstract_words
words <- words %>% filter(!word %in% c("data","using", "used", "study","area", "results","areas", "can", 
                                       "high", "along", "two", "also", "information", "due", "elsevier",
                                       "based", "may", "rights", "use", "reserved", "large", "important", 
                                       "within", ""))
wordcloud(words$word, words$freq, min.freq=500, 
          max.words=100, random.order=F, vfont=c("sans serif","plain"), 
          colors=rev(c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", 
                       "#D55E00", "#CC79A7", "#999999")))
#pie(rep(1:8), col=c("#E69F00", "#56B4E9", "#009E73", "#F0E442", "#0072B2", "#D55E00", "#CC79A7", "#999999"))
# From Okabe and Ito 2008
```

**Fig. 3.** Word cloud of the 100 most frequently used words in abstracts and titles of the considered coastal literature.  
Most studies focused on land-use and land cover change and conservation effectivness (390 and 288 publications). Species distributions and abundances and agricultural and aquacultural expansion were also studied by more than 100 studies. A fair number of studies also studied species movements and life stages, protected areas and ecosystem services, while only a marginal number of studies dealt with rapid response, ecosystem processes and climate effects (Fig. 4).  
  
<!-- Add maps of bar charts to Supplement -->

```{r topic, fig.width=10, fig.height=6}
# Identify papers related to biodiversity and conservation
d$`Species distributions and abundances`[grepl(x=d$AB, pattern="distribution") & grepl(x=d$AB, pattern="species") | grepl(x=d$AB, pattern="abundance") & grepl(x=d$AB, pattern="species") |
                  grepl(x=d$TI, pattern="distribution") & grepl(x=d$TI, pattern="species") | grepl(x=d$TI,
                                                                                                   pattern="abundance") & grepl(x=d$TI, pattern="species")] <- 1
d$`Species movements and life stages`[grepl(x=d$AB, pattern="movement") & grepl(x=d$AB, pattern="species") | grepl(x=d$AB, pattern="stage") & grepl(x=d$AB, pattern="life") |
                  grepl(x=d$TI, pattern="movement") & grepl(x=d$TI, pattern="species") | 
                    grepl(x=d$TI, pattern="life") & grepl(x=d$TI, pattern="stage")] <- 1
d$`Ecosystem processes`[grepl(x=d$AB, pattern="ecosystem process") | 
                 grepl(x=d$TI, pattern="ecosystem process")] <- 1
d$`Climate effects`[grepl(x=d$AB, pattern="climate effect") | 
                 grepl(x=d$TI, pattern="climate effect")] <- 1
d$`Rapid response`[grepl(x=d$AB, pattern="rapid response") | 
                 grepl(x=d$TI, pattern="rapid response")] <- 1
d$`Protected areas`[grepl(x=d$AB, pattern="protected area") | 
                 grepl(x=d$TI, pattern="protected area")] <- 1
d$`Ecosystem services`[grepl(x=d$AB, pattern="ecosystem service") | 
                 grepl(x=d$TI, pattern="ecosystem service")] <- 1
d$`Conservation effectivness`[grepl(x=d$AB, pattern="conservation") | 
                 grepl(x=d$TI, pattern="conservation")] <- 1
d$`Agricultural and aquacultural expansion`[grepl(x=d$AB, pattern="agriculture") | 
                 grepl(x=d$TI, pattern="agriculture") | grepl(x=d$AB, pattern="aquaculture") | 
                 grepl(x=d$TI, pattern="aquaculture")] <- 1
d$`Change in LULC and degradation and disturbance regimes`[grepl(x=d$AB, pattern="land use") | 
                 grepl(x=d$TI, pattern="land use") | grepl(x=d$AB, pattern="degradation") | 
                 grepl(x=d$TI, pattern="degradation") | grepl(x=d$AB, pattern="disturbance") | 
                 grepl(x=d$TI, pattern="disturbance")] <- 1
# 10 topics from Rose et al. 2014

d %>% select(`Species distributions and abundances`, `Species movements and life stages`, 
             `Ecosystem processes`, `Climate effects`, `Rapid response`, `Protected areas`, 
             `Ecosystem services`, `Conservation effectivness`,
             `Agricultural and aquacultural expansion`, `Change in LULC and degradation and disturbance regimes`) %>% summarise_all(sum, na.rm=T) %>% tidyr::gather(var, value) %>% 
  ggplot(aes(x=var, y=value)) + geom_bar(stat="identity", fill="#56B4E9") + theme_bw() + 
  geom_text(aes(label= value), hjust=-0.5) + 
  labs(x="", y="Number of Publications") + coord_flip() + 
  scale_y_continuous(limits=c(0,425), expand=c(0,0))
```
  
**Fig. 4.** Number of publications by research topic.  
  
<!-- Highlight number of studies that also cover another topic! -->
<!-- Check studies with two topics, themes, relevance, ... manually!!! -->

Mangrove forests were the coastal ecosystem that has been studied most using remote sensing (316 publications), followed by wetlands and coral reefs, which both have more than 200 publications. Salt marshes, intertidal zones and seagrass meadows have also been intensively studied (> 100 publications), while estuaries, kelp forests, mud flats, rocky shores and sandy shores have only been of minor interest (Fig. 5a).  

Fishes are by far the most studied taxonomic group (371 publications), followed by birds (144 publications). Invertebrates and mammals were not so frequently studied, while marine reptiles have been hardly studied at all (Fig. 5b).  

```{r eco_taxon, fig.width=10, fig.height=8}
# Coral could also be deep-sea corals, how to tease apart???
# Wetlands also does not have unique identifier
d$`Seagrass meadows`[grepl(x=d$AB, pattern="seagrass") | grepl(x=d$TI, pattern="seagrass") | 
               grepl(x=d$AB, pattern="sea grass") | grepl(x=d$TI, pattern="sea grass")] <- 1
d$`Mangroves`[grepl(x=d$AB, pattern="mangrove") | grepl(x=d$TI, pattern="mangrove")] <- 1
d$`Coral reefs`[grepl(x=d$AB, pattern="coral") | grepl(x=d$TI, pattern="coral")] <- 1
d$`Salt marshes`[grepl(x=d$AB, pattern="salt marsh") | grepl(x=d$TI, pattern="salt marsh") |
                 grepl(x=d$AB, pattern="saltmarsh") | grepl(x=d$TI, pattern="saltmarsh")] <- 1
d$`Rocky shores`[grepl(x=d$AB, pattern="rocky shore") | grepl(x=d$TI, pattern="rocky shore")] <- 1
d$`Sandy shores`[grepl(x=d$AB, pattern="sandy shore") | grepl(x=d$TI, pattern="sandy shore")] <- 1
d$`Mudflats`[grepl(x=d$AB, pattern="mud flat") | grepl(x=d$TI, pattern="mud flat") 
             | grepl(x=d$AB, pattern="mudflat") | grepl(x=d$TI, pattern="mudflat")] <- 1
d$`Intertidal zones`[grepl(x=d$AB, pattern="intertidal") | grepl(x=d$TI, pattern="intertidal")] <- 1
d$`Estuaries`[grepl(x=d$AB, pattern="estuar") | grepl(x=d$TI, pattern="estuar") | 
              grepl(x=d$AB, pattern="river") | grepl(x=d$TI, pattern="river")] <- 1
d$`Kelp forests`[grepl(x=d$AB, pattern="kelp") | grepl(x=d$TI, pattern="kelp")] <- 1
d$`Wetlands`[grepl(x=d$AB, pattern="wetland") | grepl(x=d$TI, pattern="wetland") |
               grepl(x=d$AB, pattern="wetlands") | grepl(x=d$TI, pattern="wetlands") | 
               grepl(x=d$AB, pattern="wet land") | grepl(x=d$TI, pattern="wet land") | 
               grepl(x=d$AB, pattern="wet lands") | grepl(x=d$TI, pattern="wet lands")] <- 1

p1 <- d %>% select(`Seagrass meadows`, Mangroves, `Coral reefs`, `Salt marshes`, 
                   `Rocky shores`, `Sandy shores`, `Mudflats`, `Intertidal zones`, 
                   `Estuaries`, `Kelp forests`, `Wetlands`) %>% 
  summarise_all(sum, na.rm=T) %>% tidyr::gather(var, value) %>% 
  ggplot(aes(x=var, y=value)) + geom_bar(stat="identity", fill="#56B4E9") + theme_bw() + 
  labs(x="", y="Number of Publications") + 
  geom_text(aes(label= value), hjust=-0.5) + 
  scale_y_continuous(limits=c(0,400), expand=c(0,0)) + coord_flip() + ggtitle("a)")
  #ggplot2::annotate("text", label = "a)", x = 0.75*9/5, y = 400*0.95, 
  #                  size = 6, colour = "black")

d$Fishes[grepl(x=d$AB, pattern="fish") | grepl(x=d$TI, pattern="fish")] <- 1
d$Invertebrates[grepl(x=d$AB, pattern="invertebrate") | grepl(x=d$TI, pattern="invertebrate")] <- 1 
d$Mammals[grepl(x=d$TI, pattern="mammal") | grepl(x=d$AB, pattern="mammal")] <- 1
d$Birds[grepl(x=d$AB, pattern="bird") | grepl(x=d$TI, pattern="bird")] <- 1
d$Reptiles[grepl(x=d$AB, pattern="reptile") | grepl(x=d$TI, pattern="reptile")] <- 1

p2 <- d %>% select(Fishes, Invertebrates, Mammals, Birds, Reptiles) %>% 
  summarise_all(sum, na.rm=T) %>% tidyr::gather(var, value) %>% 
  ggplot(aes(x=var, y=value)) + geom_bar(stat="identity", fill="#56B4E9") + theme_bw() + 
  labs(x="", y="Number of Publications") + 
  geom_text(aes(label= value), hjust=-0.5) + coord_flip() + 
  scale_y_continuous(limits=c(0,400), expand=c(0,0)) + ggtitle("b)")
  #ggplot2::annotate("text", label = "b)", x = 0.75, y = 400*0.95, size = 6, colour = "black")
p1 + p2 + plot_layout(ncol=1, heights=c(3,2))
```
  
**Fig. 5.** Number of studies per coastal ecosystem (a) and taxonomic group (b).  
  
<!-- Crustaceans included in invertebrates, corals? Plants? -->

By far, the most number of studies (1240) were conducted using multispectral satellite imagery. Radar data was used in 548 publications, while hyperspectral and lidar data were used the least frequent (144 and 141 publications, Fig. 6a). MODIS, Landsat and Aster data has been used the most, but Envisat, Seawifs, Sentinel and Spot data have also been used frequently (Fig. 6b).

```{r spectrum, fig.width=10, fig.height=8}
d$multispectral[grepl(x=d$AB, pattern="multispectral") | grepl(x=d$TI, pattern="multispectral") | 
                  grepl(x=d$AB, pattern="modis") | grepl(x=d$TI, pattern="modis") | 
                  grepl(x=d$AB, pattern="landsat") | grepl(x=d$TI, pattern="landsat")] <- 1
d$hyperspectral[grepl(x=d$AB, pattern="hyperspectral") | grepl(x=d$TI, pattern="hyperspectral")] <- 1
d$radar[grepl(x=d$AB, pattern="radar") | grepl(x=d$TI, pattern="radar") | 
          grepl(x=d$AB, pattern="sar") | grepl(x=d$TI, pattern="sar")] <- 1
d$lidar[grepl(x=d$AB, pattern="lidar") | grepl(x=d$TI, pattern="lidar")] <- 1

# d$multiple - Add column where it is a combination of two!

p1 <- d %>% select(multispectral, hyperspectral, radar, lidar) %>% 
  summarise_all(sum, na.rm=T) %>% tidyr::gather(var, value) %>% 
  ggplot(aes(x=var, y=value)) + geom_bar(stat="identity", fill="#56B4E9") + theme_bw() + 
  labs(x="Satellite sensor", y="Number of Publications") + coord_flip() + 
  geom_text(aes(label= value), hjust=-0.5) + ggtitle("a)") + 
  #ggplot2::annotate("text", label = "a)", x = 0.75, y = 1350*0.95, size = 6, colour = "black") + 
  scale_y_continuous(limits=c(0,1350), expand=c(0,0))

d$sentinel[grepl(x=d$AB, pattern="sentinel") | grepl(x=d$TI, pattern="sentinel")] <- 1
d$landsat[grepl(x=d$AB, pattern="landsat") | grepl(x=d$TI, pattern="landsat") | 
            grepl(x=d$AB, pattern="landsat-8") | grepl(x=d$TI, pattern="landsat-8")] <- 1
d$modis[grepl(x=d$AB, pattern="modis") | grepl(x=d$AB, pattern="aqua") | grepl(x=d$AB, pattern="terra")] <- 1
d$rapideye[grepl(x=d$AB, pattern="rapid") & grepl(x=d$AB, pattern="eye")] <- 1 
d$envisat[grepl(x=d$AB, pattern="envisat") | grepl(x=d$AB, pattern="meris")] <- 1 
d$pleiades[grepl(x=d$AB, pattern="pleiades") | grepl(x=d$TI, pattern="pleiades")] <- 1 
d$spot[grepl(x=d$AB, pattern="spot") | grepl(x=d$TI, pattern="spot")] <- 1 
# Could also be hotspot

d$seawifs[grepl(x=d$AB, pattern="seawifs") | grepl(x=d$TI, pattern="seawifs")] <- 1 
d$aquarius[grepl(x=d$AB, pattern="aquarius") | grepl(x=d$TI, pattern="aquarius")] <- 1 
d$grace[grepl(x=d$AB, pattern="grace") | grepl(x=d$TI, pattern="grace")] <- 1 
d$goci[grepl(x=d$AB, pattern="goci") | grepl(x=d$TI, pattern="goci")] <- 1 
d$ikonos[grepl(x=d$AB, pattern="ikonos") | grepl(x=d$TI, pattern="ikonos")] <- 1 
d$aster[grepl(x=d$AB, pattern="aster") | grepl(x=d$TI, pattern="aster")] <- 1 
d$hyperion[grepl(x=d$AB, pattern="hyperion") | grepl(x=d$TI, pattern="hyperion")] <- 1 
d$aviris[grepl(x=d$AB, pattern="aviris") | grepl(x=d$TI, pattern="aviris")] <- 1 

# What about radar satellites, etc. (Terra-Sar-X, Lidar?, Enmap, ...)

p2 <- d %>% select(sentinel, landsat, modis, rapideye, envisat, pleiades, 
           spot, seawifs, aquarius, grace, goci, ikonos, aster) %>% 
  summarise_all(sum, na.rm=T) %>% tidyr::gather(var, value) %>% 
  ggplot(aes(x=var, y=value)) + geom_bar(stat="identity", fill="#56B4E9") + theme_bw() + 
  labs(x="Earth observation satellite", y="Number of Publications") + 
  scale_y_continuous(limits=c(0,1350), expand=c(0,0)) + coord_flip() + 
  geom_text(aes(label= value), hjust=-0.5) + ggtitle("b)")
  #ggplot2::annotate("text", label = "b)", x = 0.75*13/5, y = 1350*0.95, size = 6, colour = "black") + 
  #theme(axis.text.x=element_text(angle=90, vjust=0.5))
p1 + p2 + plot_layout(ncol=1, height=c(2,3))
```

**Fig. 6.** Number of publications per spectrum (a) and Earth-observation satellite (b).   
  
## Discussion

<!-- Check out Roll et al. 2018 Conservation Biology !!! For discussion on methodology.
Check out Murray et al. 2018  Science of the Total Environment
-->

Remote sensing has experienced an increasing use in recent decades (Pettorelli et al. 2014), which corresponds with the exponential increase in studies related to coastal remote sensing (Fig. 1). In recent years, RS has been more widely applied in the marine and coastal realm, but generally lies far behind the terrestrial application of RS. New upcoming satellite missions and open access to data products together with an increasing effort in teaching the analysis of remote sensing data to non-experts is most likely the cause for this. Nevertheless, technological difficulties are still hampering the use of RS, in particular in marine systems.  

<!-- Structure discussion according to the results section! -->
...

<!-- Add Figure on total publications, publications terrestrial, marine, coastal to Supplement!!! -->

<!--
### Research theme and topic

Species distribution and abundance is commonly measured in ecological research. Traditionally, trawling and other extractive methods have been used for determining the abundance and distribution of marine species. Advances in technology have allowed for non-destructive remote alternatives, such as camera traps, ROVs and tethered video surveys. Camera traps can not only be used to record the species composition of fish assemblages and estimate its abundance, but also to observe the behaviour of individuals. They have especially been applied in the deep sea (Bailey et al. 2007), as it is a more cost effective and simple way of obtaining ecological data at great depths. ROVs have increasingly been used to estimate fish density. However, some fish species are thought to be attracted to or avoid the light and noise emissions of ROVs (Trenkel et al. 2004). These methods are particularly useful for studying vulnerable marine ecosystems and protected areas or species (Marouchos et al. 2011, Martinez et al. 2011). Acoustic surveys offer another way of predicting the spatial distribution, species composition and length of fish species by analysing the acoustic backscatter of fish and correlating them to trawl or video data (Juntunen et al. 2012). 

Satellite measurements have enabled us to better assess and predict natural hazards. Earthquakes, floods, tsunamis, storms and fires can be readily analysed over a number of spatial and temporal scales using remote sensing (Gillespie et al. 2007). The effects of tsunamis on the landscape can be assessed using high-resolution satellite imagery (Umitsu et al. 2007) and areas of high risk can be identified for the future.  

The spatial distribution of species observations is often limited to individual point observations. Species distribution models (SDMs) provide a mean of interpolating point species abundance and distribution data and thus have a high relevance for marine conservation and management (Degraer et al. 2008). Species distribution modelling can provide an important function for managing invasive species (Bernal et al. 2015) and for stearing evidence-based conservation efforts (Guinotte and Davies 2014). However, the application of SDMs in the marine environment is still sparse, in particular in deep-sea marine ecosystems (Marshall et al. 2014).

Marine protected areas (MPAs) are an important tool for conserving vulnerable marine ecosystems, marine biodiversity and controlling anthropogenic impacts, such as fishing. Aerial imagery can be used to identify habitat categories, which are a cost-effective method for the identification of high-priority areas for sustaining marine diversity of coastal ecosystems (Ward et al. 1999). 

The analysis of animal movement patterns is further an important tool for effective management and conservation of species and habitats. For many migratory marine species, such as cetaceans, sharks and marine turtles, as well as for marine birds, the breeding as well as the foraging grounds need to be protected and both need to be linked (Pendoley et al. 2014). Therefore, the complex interaction of seasonality, life-stage and availability of food resources needs to be considered in order to come up with a successful conservation strategy (Lander et al. 2013). Radio telemetry offers a way of gathering species presence data over time, linking temporal and spatial changes in species abundance to changes in environmental conditions, and hereby provide important information for the management of habitats for threatened or endangered species (Jacoby et al. 2012).  

Unmanned aerial vehicles, such as balloons or helicopters, can be an inexpensive alternative to satellite and airborne sensors and offer a much higher spatial resolution. They are particularly useful for measuring ecosystem processes on a small scale and can be used for producing digital elevation models, the mapping of land cover or to survey large animals (Scheritz & Dietrich 2002, Koh & Wich 2012). Marine ecosystems range from complex heterogeneous habitats, such as rocky shores (Meager et al. 2011) and coral reefs (Polunin 2007), to homogeneous and featureless soft-sediment seafloors (Kraan et al. 2009). Habitat complexity is an important driver of ecosystem structure and functioning. Meager et al. (2011), for example, used high resolution 3D fractal surface and temperature measurements to analyse the effect of habitat complexity of intertidal rocky shores on species abundance and body size of invertebrates. Data derived from airborne digital cameras further offers a way to count and map individual animals, for example of marine birds, in order to estimate population size and animal-habitat interactions. The resulting knowledge can for example be applied in environmental impact assessments of marine installations, such as offshore wind farms (Groom et al. 2013). Sea-ice extent and thickness are important climate variables in the Polar Regions. According to satellite measurements summer Arctic sea‐ice extent has decreased over the past three decades. However, it is also subject to large inter-annual and regional variations (Serreze et al. 2007, Stroeve et al. 2012). Satellite measurements further show that the decline in sea-ice extent has been accompanied by a decline in sea-ice volume (Laxon et al. 2013). Using an electromagnetic-induction (EM) system sea ice thickness can be measured from icebreakers and small helicopters (Haas et al. 2009). Using the EM system, Haas et al. (2010) found that modal thickness of old ice in the Arctic Ocean had changed little from 2007 to 2009, and remained within the expected range of natural variability.  

Marine systems provide the majority of global ecosystem services, with most of it coming from coastal systems (Costanza et al. 1997). Seagrass meadows, for example, provide a great economic and ecological value of goods and services, such as nutrient cycling (McGlathery et al. 2007), enhancement of fish productivity (Watson et al. 1993), as well as habitat and food provisioning (de la Torre-Castro & Rönnbäck 2004). Yet, seagrasses have been disappearing since they were initially recorded in 1879 and rates of decline have even accelerated in the recent past (Waycott et al. 2009). Other important marine ecosystems, such as mangroves (Valiela et al. 2001) and coral reefs (Hoegh-Guldberg et al. 2007), show similar losses. Remote sensing offers a way of examining the state and extent of these ecosystems on a global scale (Giri et al. 2011). Mangroves have shown resilience to disturbances caused by shoreline changes and are likely to be resilient to anthropogenically amplified sea level rise in the future, while climate change is thought to cause a global loss of 10-15% of mangrove forests. However, anthropogenic disturbances, such as deforestation, give rise to much greater concern, as the annual deforestation rate of mangroves is 12 % (Alongi 2008). Remotely sensed imagery can be used for tracking the deforestation of mangroves (Rahman et al. 2013) and can further be used to assess the mangrove development, within mixed mangrove-aquaculture farming systems, which helps the government to regulate the extent of aquaculture farming areas (Vo et al. 2013) and thereby limit deforestation. Ocean acidification will have a tremendous effect on coral species in the near future (Fabry et al. 2008). In a worst-case scenario, the majority of cold-water coral reefs is expected to be exposed to corrosive waters by 2060 (Jackson et al. 2014). In order to manage coral reefs in an era of changing climate, we need to be able to measure and map coral reef resilience. Coral and algal cover, as well as a range of coral reef stressors, can be estimated using RS (Knudby et al. 2014). Bleached corals further have a strong optical signal, which can also be detected from airborne or satellite sensors (Andréfouët et al. 2002). This can then be used to establish marine protected areas that harbour coral reefs, which are resilient to climate change.  

### Ecosystems & taxonomic group

Models suggest that mangroves forests of at least 100 m in width can significantly reduce the wave flow pressure of tsunamis (Alongi 2008). RS of the spatial distribution of mangrove forests could thus also help to identify areas that are particularly vulnerable to tsunamis and could so help to reduce the risk of tsunamis by informed future development decisions and restoration projects.  

SDMs, for example, have been widely used, but are still lacking for the deep-sea and on a regional scale. The movement, in particular the foraging behaviour, of marine animals is largely unknown, as tagging of animals is much more difficult than for terrestrial species. The effects of environmental changes due to natural and anthropogenic causes on marine ecosystems have been widely studied, but there is still a huge potential in increasing our knowledge by using RS.  

For bird species, such as the Manx shearwater (*Puffinus puffinus*), breeding colonies are often designated as special protection areas, but they hardly ever extent into the marine environment. Manx shearwaters form rafts and then spent a considerable time at sea. Using radio-telemetry the rafting distribution of Manx shearwaters was investigated and the maximum extent of rafting ranges was identified. This could then be used to define protected areas for birds in the marine realm (Wilson et al. 2009).  

### Technological advancements & drawbacks

Mention paper by Schulte-to-Bühne et al. 2017

Sea-ice extent and thickness are important climate variables in the Polar Regions. According to satellite measurements summer Arctic sea‐ice extent has decreased over the past three decades. However, it is also subject to large inter-annual and regional variations (Serreze et al. 2007, Stroeve et al. 2012). Satellite measurements further show that the decline in sea-ice extent has been accompanied by a decline in sea-ice volume (Laxon et al. 2013). Using an electromagnetic-induction (EM) system sea ice thickness can be measured from icebreakers and small helicopters (Haas et al. 2009).  

RS has the advantage of providing continuous spatio-temporal data. However, different data products only occur on a certain spatial or temporal scale, especially long-term data products. Furthermore, temporal and spatial resolution are conflicting each other, as a high temporal resolution is usually sacrificed by a lower spatial resolution or a smaller spatial extent, and vice versa. Optical sensors, in particular of satellites, cannot pass through clouds. Thus, areas with persistent cloud cover usually lack data (Turner et al. 2003), limiting the use of RS in areas with high cloud cover, such as the tropics.  

Furthermore, remote sensing products are usually based on a large number of complex algorithms, which require baseline in situ data from ecological surveys in order to ground-truth our estimates. These estimates can have huge associated errors, for example Dierssen (2010) argues that chlorophyll α estimates from satellites can be in error by a factor of 5 or more. Estimation uncertainty and error is further likely to increase in the future, with an increasing variation in future climatic conditions.  

Despite these drawbacks, RS is a powerful indispensable tool for ecological research and marine research in particular. Future high-resolution satellites and more readily-available remotely sensed data products will help to increase our scientific understanding of marine communities even further and allow for the implementation of these data sources in conservation management and policy-decision making.  

## Additional Resources

Earth observation from satellites offers vast potential for fisheries applications, including management of marine resources, stock assessment, marine aquaculture, and fish harvesting. - (Stuart et al. 2011)/IOCCG, 2009 As outlined in the International Ocean Colour Coordinating Group (IOCCG) report, there is a broad range of fisheries-related applications of remotely sensed data, including bycatch reduction, detection of harmful algal blooms, aquaculture site selection, identifying marine managed areas, and describing habitat changes. Satellite data can be used to investigate fluctuations in food availability during the critical period of larval stages and can be used to infer information about the year classes of exploited stocks (Stuart et al. 2011).

Californian squid fishery targets nearshore spawning aggregations, which raises concern with regards to its sustainability, as reproductive activities and subsequent egg-laying might occur at much slower rates. High-resolution acoustic surveys using side scan sonar showed that squids generally prefer sandy habitats as their primary egg mop sites, however the depth distribution differed significantly between surveys and spatial distribution differed across years. This highlights the use of side scan sonar surveys as a tool to implement sustainable management of squid fisheries through improved monitoring, designation and adaptive management of no-take areas and stock assessments (Young et al. 2011).
Sardine recruitment in the California Current is influenced by the Pacific Decadal Oscillation (PDO) index (Zwolinski and Demer 2014).

Vessel monitoring system (VMS) data provide information on fishing activity on large spatial scales (Lambert et al. 2012/Ref?), however government regulations limit the use of potentially available data (Hinz et al. 2013). Fishing vessel identification from satellite images further provides a complementary measure to control fishing activities. Vessel detection rates from satellite images were between 75-100% depending on the image type and vessel size (Kourti et al. 2005). Satellite monitoring of fishing vessel activity indicates that fisheries closures can work effectively with good compliance by international fleets even in remote areas (Hall-Spencer et al. 2009).

Campbell et al. 2013, Bastardie et al. 2010

### Acoustic sensors

Shipborne acoustic measurements can be divided into single beam and multi beam measurements. Single beam measurements use acoustic energy to collect measurements of the seafloor depth, while multi beam measurements are active measurements of acoustic energy and measure seafloor depth as well as the character of the seafloor (Roff & Zacharias 2011).
Seafloor topography not only affects ocean circulation and mixing, which influences the Earth’s climate (Kunze and Llewellyn Smith 2004; Munk and Wunsch 1998), but also has an effect on biological diversity and marine food resources (Koslow 1997). Despite the vast availability of multi beam echo sounders, most of the ocean floor still remains unexplored (Becker et al. 2009). Currently, only 6.5 % of the ocean floor has been surveyed with soundings (Original source?/Yesson et al. 2011). Bathymetry data from multi beam surveys can be used to obtain various descriptors of the seabed terrain, such as slope and aspect. These descriptors can be applied as predictor variables in species distribution models (SDMs), particularly for the deep-sea where data on the benthic fauna might be limited (Wilson et al. 2007). Yesson et al. (2011) further used global bathymetric data at 30 arc-sec resolution to identify seamounts and knolls on a global scale. Seamounts and knolls are undersea mountains, which are thought to provide an important habitat for demersal deep-sea fish and benthic invertebrates (Source!!!). In total, they found 33452 seamounts and 138412 knolls. Comparing this to known occurrences of seamounts, they were able to find 94 % of seamounts, although some areas along ridges and seafloor spreading areas had an overestimation of seamount numbers.
--->

## References

<!-- Update references according to journal guideline, check all cited papers appear in References and if all references are cited in the text. -->

Ahumada JA, Silva CEF, Gajapersad K, Hallam C, Hurtado J, Martin E, McWilliam A, Mugerwa B, O’Brien T, Rovero F, Sheil D, Spironello WR, Winarni N, Andelman SJ (2011) Community structure and diversity of tropical forest mammals: data from a global camera trap network. Philos Trans R Soc Lond B Biol Sci 366:2703–2711  

Alongi DM (2008) Mangrove forests: Resilience , protection from tsunamis , and responses to global climate change. 76:1–13  

Andréfouët S, Berkelmans R, Odriozola L, Done T, Oliver J, Müller-Karger F (2002) Choosing the appropriate spatial resolution for monitoring coral bleaching events using remote sensing. Coral Reefs Volume 21:147–154  

Bailey DM, King NJ, Priede IG (2007) Cameras and carcasses: historical and current methods for using artificial food falls to study deep-water animals. 350:179–191  

Baldock J, Bancroft KP, Williams M, Shedrawi G, Field S (2014) Accurately estimating local water temperature from remotely sensed satellite sea surface temperature: A near real-time monitoring tool for marine protected areas. Ocean Coast Manag 96:73–81  

<!--
Bastardie F, Nielsen JR, Ulrich C, Egekvist J, Degel H (2010) Detailed mapping of fishing effort and landings by coupling fishing logbooks with satellite-recorded vessel geo-location. Fish Res 106:41–53  
-->

Becker JJ, Sandwell DT, Smith WHF, Braud J, Binder B, Depner J, Fabre D, Factor J, Ingalls S, Kim S-H, Ladner R, Marks K, Nelson S, Pharaoh a., Trimmer R, Rosenberg J Von, Wallace G, Weatherall P (2009) Global Bathymetry and Elevation Data at 30 Arc Seconds Resolution: SRTM30_PLUS. Mar Geod 32:355–371  

Belluco E, Camuffo M, Ferrari S, Modenese L, Silvestri S, Marani A, Marani M (2006) Mapping salt-marsh vegetation by multispectral and hyperspectral remote sensing. Remote Sens Environ 105:54–67  

Bernal MA, Sinai NL, Rocha C, Gaither MR, Dunker F, Rocha LA (2015) Long-term sperm storage in the brownbanded bamboo shark Chiloscyllium punctatum. J Fish Biol 86:1171–1176  

Biber MF, Duineveld GCA, Lavaleye MSS, Davies AJ, Bergman MJN, Beld IMJ van den (2014) Investigating the association of fish abundance and biomass with cold-water corals in the deep Northeast Atlantic Ocean using a generalised linear modelling approach. Deep Res Part II Top Stud Oceanogr 99:134–145  

Bowers DG, Hill PS, Braithwaite KM (2014) The effect of particulate organic content on the remote sensing of marine suspended sediments. Remote Sens Environ 144:172–178  

Brewington L, Frizzelle BG, Walsh SJ, Mena CF, Sampedro C (2014) Remote Sensing of the Marine Environment: Challenges and Opportunities in the Galapagos Islands of Ecuador. In: Denkinger J, Vinueza L (eds) The Galapagos Marine Reserve: A Dynamic Social-Ecological System.p 322  

Chen LC, Rau JY (1998) Detection of shoreline changes for tideland areas using multi-temporal satellite images. Int J Remote Sens 19:3383–3397  

Costanza R, D’Arge R, Groot R de, Farber S, Grasso M, Hannon B, Limburg K, Naeem S, O’Neill R, Paruelo J, Raskin R, Sutton P, Belt M van den (1997) The value of the world’s ecosystem services and natural capital. Nature 387  

Cousins NJ, Linley TD, Jamieson AJ, Bagley PM, Blades H, Box T, Chambers R, Ford A, Shields MA, Priede IG (2013) Bathyal demersal fishes of Charlie-Gibbs Fracture Zone region (49-54°N) of the Mid-Atlantic Ridge: II. Baited camera lander observations. Deep Res Part II Top Stud Oceanogr 98:397–406  

Deguignet M, Juffe-Bignoli D, MacSharry B, Burgess ND, Kingston N (2014) 2014 United Nations List of Protected Areas. UNEP-WCMC, Cambridge, UK  

Dierssen HM (2010) Perspectives on empirical approaches for ocean color remote sensing of chlorophyllin a changing climate. Proc Natl Acad Sci 107:17073–17078  
  
Edgar GJ, Stuart-Smith RD, Willis TJ, Kininmonth S, Baker SC, Banks S, Barrett NS, Becerro MA, Bernard ATF, Berkhout J, Buxton CD, Campbell SJ, Cooper AT, Davey M, Edgar SC, Försterra G, Galván DE, Irigoyen AJ, Kushner DJ, Moura R, Parnell PE, Shears NT, Soler G, Strain EM a, Thomson RJ (2014) Global conservation outcomes depend on marine protected areas with five key features. Nature 506:216–20  

Eriksen CC, Osse TJ, Light RD, Wen T, Lehman TW, Sabin PL, Ballard JW, Chiodi AM (2001) Seaglider: A Long-Range Autonomous Underwater Vehicle for Oceanographic Research. 26:424–436  

Evans RD, Murray KL, Field SN, Moore JAY, Shedrawi G, Huntley BG, Fearns P, Broomhall M, Mckinna LIW, Marrable D (2012) Digitise This! A Quick and Easy Remote Sensing Method to Monitor the Daily Extent of Dredge Plumes. 7  

Fabry VJ, Seibel BA, Feely RA, Orr JC (2008) Impacts of ocean acidification on marine fauna and ecosystem processes. ICES J Mar Sci 65:414–432  

Game ET, Grantham HS, Hobday AJ, Pressey RL, Lombard AT, Beckley LE, Gjerde K, Bustamante R, Possingham HP, Richardson AJ (2009) Pelagic protected areas: the missing dimension in ocean conservation. Trends Ecol Evol 24:360–369  

Gillespie TW, Chu J, Frankenberg E, Thomas D (2007) Assessment and prediction of natural hazards from satellite imagery. Prog Phys Geogr 31:459–470  

Giri C, Ochieng E, Tieszen LL, Zhu Z, Singh A, Loveland T, Masek J, Duke N (2011) Status and distribution of mangrove forests of the world using earth. ... :154–159  

Groom G, Stjernholm M, Nielsen RD, Fleetwood A, Petersen IK (2013) Remote sensing image data and automated analysis to describe marine bird distributions and abundances. Ecol Inform 14:2–8  

Guinotte JM, Davies AJ (2014) Predicted Deep-Sea Coral Habitat Suitability for the U.S. West Coast. 9  

Haas C, Lobach J, Hendricks S, Rabenstein L, Pfaffling A (2009) Helicopter-borne measurements of sea ice thickness, using a small and lightweight, digital EM system. J Appl Geophys 67:234–241  

Hoegh-Guldberg O, Mumby PJ, Hooten a J, Steneck RS, Greenfield P, Gomez E, Harvell CD, Sale PF, Edwards a J, Caldeira K, Knowlton N, Eakin CM, Iglesias-Prieto R, Muthiga N, Bradbury RH, Dubi a, Hatziolos ME (2007) Coral reefs under rapid climate change and ocean acidification. Science 318:1737–1742  

Huvenne VAI, Tyler PA, Masson DG, Fisher EH, Hauton C, Hühnerbach V, Bas TP, Wolff G a (2011) A picture on the wall: Innovative mapping reveals cold-water coral refuge in submarine canyon. PLoS One 6:e28755  

Jackson E, Davies A, Howell K, Kershaw P, Hall-Spencer J (2014) Future-proofing marine protected area networks for cold water coral reefs. ICES J Mar Sci  

Jacoby DMP, Brooks EJ, Croft DP, Sims DW (2012) Developing a deeper understanding of animal movements and spatial dynamics through novel application of network analyses. :574–583  

Jeffreys RM, Lavaleye MSS, Bergman MJN, Duineveld GCA, Witbaard R, Linley T (2010) Deep-sea macrourid fishes scavenge on plant material: Evidence from in situ observations. Deep Sea Res Part I Oceanogr Res Pap 57:621–627  

Juntunen T, Vanhatalo J, Peltonen H, Mäntyniemi S (2012) Bayesian spatial multispecies modelling to assess pelagic fish stocks from acoustic- and trawl-survey data. ICES J Mar Sci 69:95–104  

Knudby A, Pittman SJ, Maina J, Rowlands G (2014) Remote Sensing and Modeling of Coral Reef Resilience. In: Remote Sensing and Modeling: Advances in Coastal and Marine Resources.p 494  

Koh LP, Wich SA (2012) Dawn of drone ecology: low-cost autonomous aerial vehicles for conservation. 5:121–132

Kraan C, Meer J Van Der, Dekinga A, Piersma T (2009) Patchiness of macrobenthic invertebrates in homogenized intertidal habitats: hidden spatial structure at a landscape scale. 383:211–224  

la Torre-Castro M de, Rönnbäck P (2004) Links between humans and seagrasses—an example from tropical East Africa. Ocean Coast Manag 47:361–387  

Lagerloef G, Colomb FR, Vine D Le, Wentz F, Yueh S, Ruf C, Lilly J, Gunn J, Chao Y, DeCharon A, Feldman G, Swift C (2008) The Aquarius/SAC-D Mission: Designed to Meet the Salinity Remote-Sensing Challenge. Oceanography 21:68–81  

Lagerloef GSE, Swift CT, LeVine DM (1995) Sea Surface Salinity: The Next Remote Sensing Challenge. Oceanography 8:44–50  

Lander ME, Fritz LW, Johnson DS, Logsdon MG (2013) Population trends of Steller sea lions (Eumetopias jubatus) with respect to remote sensing measures of chlorophyll-a in critical habitat. Mar Biol 160:195–209  

Laxon SW, Giles K a., Ridout AL, Wingham DJ, Willatt R, Cullen R, Kwok R, Schweiger A, Zhang J, Haas C, Hendricks S, Krishfield R, Kurtz N, Farrell S, Davidson M (2013) CryoSat-2 estimates of Arctic sea ice thickness and volume. Geophys Res Lett 40:732–737  

Leifer I, Lehr WJ, Simecek-Beatty D, Bradley E, Clark R, Dennison P, Hu Y, Matheson S, Jones CE, Holt B, Reif M, Roberts DA, Svejkovsky J, Swayze G, Wozencraft J (2012) State of the art satellite and airborne marine oil spill remote sensing: Application to the BP Deepwater Horizon oil spill. Remote Sens Environ 124:185–209  

Li R, Liu J-K, Felus Y (2001) Spatial Modeling and Analysis for Shoreline Change Detection and Coastal Erosion Monitoring. Mar Geod 24:1–12  

Lirman D, Gracias NR, Gintert BE, Gleason ACR, Reid RP, Negahdaripour S, Kramer P (2007) Development and application of a video-mosaic survey technology to document the status of coral reef communities. Environ Monit Assess 125:59–73  

Lorance P, Trenkel V (2006) Variability in natural behaviour, and observed reactions to an ROV, by mid-slope fish species. J Exp Mar Bio Ecol 332:106–119  

Marouchos A, Sherlock M, Barker B, Williams A (2011) Development of a stereo deepwater Baited Remote Underwater Video System (DeepBRUVS). Ocean 2011 IEEE - Spain:1–5  
Marshall CE, Glegg GA, Howell KL (2014) Species distribution modelling to support marine conservation planning: The next steps. Mar Policy 45:330–332  

Martinez I, Jones EG, Davie SL, Neat FC, Wigham BD, Priede IG (2011) Variability in behaviour of four fish species attracted to baited underwater cameras in the North Sea. Hydrobiologia 670:23–34  

McGlathery KJ, Sundbäck K, Anderson IC (2007) Eutrophication in shallow coastal bays and lagoons: The role of plants in the coastal filter. Mar Ecol Prog Ser 348:1–18  

Meager J, Schlacher T, Green M (2011) Topographic complexity and landscape temperature patterns create a dynamic habitat structure on a rocky intertidal shore. Mar Ecol Prog Ser 428:1–12  

<!--
Murphy RJ, Underwood AJ, Tolhurst TJ, Chapman MG (2008) Remote Sensing of Environment Field-based remote-sensing for experimental intertidal ecology: Case studies using hyperspatial and hyperspectral data for New South Wales ( Australia ). 112:3353–3365  
-->
Nagendra H, Lucas R, Honrado JP, Jongman RHG, Tarantino C, Adamo M, Mairota P (2013) Remote sensing for conservation monitoring: Assessing protected areas, habitat extent, habitat condition, species diversity, and threats. Ecol Indic 33:45–59  

Ozesmi SL, Bauer ME (2002) Satellite remote sensing of wetlands. Wetl Ecol Manag 10:381–402  

Palmer SCJ, Kutser T, Hunter PD (2014) Remote sensing of inland waters: Challenges , progress and future directions. Remote Sens Environ 

Parijs SM Van, Clark CW, Sousa-Lima RS, Parks SE, Rankin S, Risch D, Opzeeland IC Van (2009) Management and research applications of real-time and archival passive acoustic sensors over varying temporal and spatial scales. Mar Ecol Prog Ser 395:21–36  

Pasqualini V, Pergent-Martini C, Clabaut P, Marteel H, Pergent G (2001) Integration of Aerial Remote Sensing, Photogrammetry, and GIS Technologies in Seagrass Mapping. Photogramm Eng Remote Sensing 67:99–105  

Pendoley KL, Schofield G, Whittock PA, Ierodiaconou D, Hays GC (2014) Protected species use of a coastal marine migratory corridor connecting marine protected areas. :1455–1466  

Pereira HM, Leadley PW, Proença V, Alkemade R, Scharlemann JPW, Fernandez-Manjarrés JF, Araújo MB, Balvanera P, Biggs R, Cheung WWL, Chini L, Cooper HD, Gilman EL, Guénette S, Hurtt GC, Huntington HP, Mace GM, Oberdorff T, Revenga C, Rodrigues P, Scholes RJ, Sumaila UR, Walpole M (2010) Scenarios for global biodiversity in the 21st century. Sciencexpress 330:1496–1501  

Pettorelli N, Laurance WF, Brien TGO, Wegmann M, Nagendra H, Turner W (2014) Satellite remote sensing for applied ecologists: opportunities and challenges. ... :839–848  

Polunin SKWNAJGNVC (2007) Appraisal of visual assessments of habitat complexity and benthic composition on coral reefs. ... :1069–1076  

Rahman AF, Dragoni D, Didan K, Barreto-Munoz A, Hutabarat J a. (2013) Detecting large scale conversion of mangroves to aquaculture with change point and mixed-pixel analyses of high-fidelity MODIS data. Remote Sens Environ 130:96–107  

Raitsos DE, Pradhan Y, Brewin RJW, Stenchikov G, Hoteit I (2013) Remote Sensing the Phytoplankton Seasonal Succession of the Red Sea. 8  

R Core Team (2020). R: A language and environment for statistical computing. R Foundation for Statistical Computing, Vienna, Austria. URL https://www.R-project.org/.  

Römer H, Willroth P, Kaiser G, Vafeidis a. T, Ludwig R, Sterr H, Revilla Diez J (2012) Potential of remote sensing techniques for tsunami hazard and vulnerability analysis-a case study from Phang-Nga province, Thailand. Nat Hazards Earth Syst Sci 12:2103–2126  
Satyanarayana B, Mohamad KA, Idris IF, Husain M-L, Dahdouh-Guebas F (2011) Assessment of mangrove vegetation based on remote sensing and ground-truth measurements at Tumpat, Kelantan Delta, East Coast of Peninsular Malaysia. Int J Remote Sens 32:1635–1650  

Scheritz M, Dietrich R (2002) Digital Elevation Model of Polygonal Patterned Ground on Samoylov Island , Siberia , Using Small-Format Aerial Photography. :1589–1594  

Serreze MC, Holland M, Stroeve JC (2007) Perspectives on the Arctic’s shrinking sea-ice cover. Science (80- ) 315:1533–1536  

Smit AJ, Roberts M, Anderson RJ, Dufois F, Dudley SFJ (2013) A Coastal Seawater Temperature Dataset for Biogeographical Studies: Large Biases between In Situ and Remotely-Sensed Data Sets around the Coast of South Africa. 8  

Smith WH (1997) Global Sea Floor Topography from Satellite Altimetry and Ship Depth Soundings. Science (80- ) 277:1956–1962  

Stroeve JC, Serreze MC, Holland MM, Kay JE, Malanik J, Barrett AP (2012) The Arctic’s rapidly shrinking sea ice cover: A research synthesis. Clim Change 110:1005–1027  

Stuart V, Platt T, Sathyendranath S (2011) The future of fisheries science in management: a remote-sensing perspective.  

Trenkel V, Lorance P, Mahevas S (2004) Do visual transects provide true population density estimates for deepwater fish? ICES J Mar Sci 61:1050–1056  

Umitsu M, Tanavud C, Patanakanog B (2007) Effects of landforms on tsunami flow in the plains of Banda Aceh, Indonesia, and Nam Khem, Thailand. Mar Geol 242:141–153  

Valiela I, Bowen JL, York JK (2001) Mangrove Forests: One of the World’s Threatened Major Tropical Environments. Bioscience 51:807  

Vo QT, Oppelt N, Leinenkugel P, Kuenzer C (2013) Remote sensing in mapping mangrove ecosystems - an object-based approach. Remote Sens 5:183–201  

Ward TJ, Vanderklift MA, Nicholls AO, Kenchington RA (1999) Selecting Marine Reserves Using Habitats and Species Assemblages As Surrogates For Biological Diversity. Ecol Appl 9:691–698  

Watson R a., Coles RG, Long WJL (1993) Simulation estimates of annual yield and landed value for commercial penaeid prawns from a tropical seagrass habitat, northern Queensland, Australia. Aust J Mar Freshw Res 44:211–219  

Waycott M, Duarte CM, Carruthers TJB, Orth RJ, Dennison WC, Olyarnik S, Calladine A, Fourqurean JW, Heck KL, Hughes a R, Kendrick G a, Kenworthy WJ, Short FT, Williams SL (2009) Accelerating loss of seagrasses across the globe threatens coastal ecosystems. Proc Natl Acad Sci U S A 106:12377–81  

Wilson LJ, Mcsorley CA, Gray CM, Dean BJ, Dunn TE, Webb A, Reid JB (2009) Radio-telemetry as a tool to define protected areas for seabirds in the marine environment. Biol Conserv 142:1808–1817  

Yoerger D, Bradley A, Jakuba M, German C, Shank T, Tivey M (2007) Autonomous and Remotely Operated Vehicle Technology for Hydrothermal Vent Discovery, Exploration, and Sampling. Oceanography 20:152–161  

## Supplementary Material

### Supplementary Tables

**Table S1.** Search terms used and the resulting number of publications from the Web of Science Core Collection.

Search term  | Number of publications
------------------ | -----------------------
remote sensing* coast | 1960
remote sensing* marine | 2268
landsat* coast | 516
landsat* marine | 248
MODIS* coast | 535
MODIS* marine | 570
Sentinel* coast | 282
Sentinel* marine | 547
remote sensing* seagrass | 228
remote sensing* corals | 552
remote sensing* mangroves | 472
remote sensing* salt marsh | 203
remote sensing* salt marshes | 203
remote sensing* biodiversity | 1812
remote sensing* conservation | 2658
   
   
  
**Table S2.** Category, value and the keywords used to identifiy the value of each category for each publication. All keywords word searched for in the title and abstract of each publication.  

Category | Value  | Keywords
------------------ | ------------------ | -----------------------
relevance | terrestrial | land, terrestrial
relevance | coastal | coast, estuar, intertidal, shore
relevance  | marine | marine, deep-sea
country | countries | countries
theme | biodiversity | biodiversity
theme  | conservation | conservation
theme  | climate change | climate change
theme  | pollution | pollution
topic | Species distributions and abundances | distribution & species, abundance & species
topic  | Species movements and life stages | movement species, life & stage
topic  | Ecosystem processes | ecosystem process
topic  | Climate effects | climate effect
topic  | Rapid response | rapid response
topic  | Protected areas | protected area
topic  | Ecosystem services | ecosystem service
topic  | Conservation effectivness | conservation
topic  | Agricultural and aquacultural expansion | agriculture, aquaculture
topic | Change in LULC and degradation and disturbance regimes | land use, degradation, disturbance
ecosystem | Seagrass meadows | sea grass, seagrass
ecosystem  | Coral reefs | coral
ecosystem  | Mangroves | mangrove
ecosystem  | Salt marshes | salt marsh, saltmarsh
ecosystem  | Rocky shores | rocky shore
ecosystem  | Sandy shores | sandy shore
ecosystem  | Mudflats | mud flat, mudflat
ecosystem  | Estuary | estuary, estuaries, river
ecosystem  | Intertidal zones | intertidal
ecosystem  | Kelp forests | kelp
taxonomic group | Fishes | fish
taxonomic group  | Mammals | mammal
taxonomic group  | Birds | bird
taxonomic group  | Invertebrates | invertebrate
taxonomic group  | Reptiles | reptile
sensor | multispectral | multispectral, modis, landsat
sensor  | hyperspectral | hyperspectral
sensor  | radar | radar, sar
sensor  | lidar | lidar
EO satellite | sentinel | sentinel
EO satellite  | landsat |  landsat-8
EO satellite  | MODIS | modis, aqua, terra
EO satellite  | RapidEye | rapid & eye
EO satellite  | Envisat | envisat, meris
EO satellite  | Pleiades | pleiades
EO satellite  | Spot | spot
EO satellite  | SeaWifs | seawifs
EO satellite  | Aquarius | aquarius
EO satellite  | Grace | grace
EO satellite  | GOCI | goci
EO satellite  | Ikonos | ikonos
EO satellite  | Aster | aster
EO satellite  | Hyperion | hyperion
EO satellite  | Aviris | aviris

### Supplementary Figures
 
```{r pub_dupl, fig.width=10, fig.height=6}
d %>% select(PY, search) %>% group_by(PY, search) %>% summarise(sum = n()) %>% 
  tidyr::drop_na() %>% ungroup() %>% filter(sum > 5) %>% 
  mutate(PY = cut(PY, breaks=c(1989, 1995, 2000, 2005, 2010, 2015, 2020))) %>%
  ggplot(aes(x=factor(search), y=sum, fill=PY)) + geom_bar(stat="identity") + 
  scale_y_continuous(limits=c(0,850), breaks=seq(0,850,by=50), expand=c(0,0)) + 
  scale_fill_manual(name="Year", values=c('#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e','#e6ab02')) + 
  labs(x="", y="Number of Publications") + 
  theme_bw() + coord_flip()
```
  
**Fig. S1.** Number of publications that were identified by one or multiple search terms. Only search term combinations that yielded more than 5 publications are shown.  

```{r pub_sm, fig.width=10, fig.height=6}
# Identify papers related to biodiversity and conservation
d$`Biodiversity`[grepl(x=d$AB, pattern="biodiversity") | 
                  grepl(x=d$TI, pattern="biodiversity")] <- 1
d$`Conservation`[grepl(x=d$AB, pattern="conservation") | 
                 grepl(x=d$TI, pattern="conservation")] <- 1
d$`Climate Change`[grepl(x=d$AB, pattern="climate change") | 
                 grepl(x=d$TI, pattern="climate change")] <- 1
d$`Pollution`[grepl(x=d$AB, pattern="pollution") | 
                 grepl(x=d$TI, pattern="pollution")] <- 1
d$Sum <- rowSums(d[,c("Biodiversity", "Conservation", "Climate Change", "Pollution")], na.rm=T)
d$None[d$Sum!=0] <- NA
d$None[d$Sum==0] <- 1
d$Multiple[d$Sum>1] <- 1
# d$multiple - Add column where it is a combination of two!

# Plot number of publications over time
d %>% select(-TI, -AU, -SO) %>%
  select(PY, Biodiversity, Conservation, `Climate Change`, Pollution, None, Multiple) %>% 
  tidyr::gather(Theme, value, -c(PY)) %>% group_by(PY, Theme) %>% tidyr::drop_na() %>% 
  summarise(sum = n()) %>% ungroup() %>% 
  tidyr::complete(PY, Theme, fill = list(sum=0)) %>% arrange(Theme, PY) %>%
  mutate(Theme = factor(Theme, levels=c("None", "Biodiversity", "Climate Change", "Conservation", "Pollution", "Multiple"))) %>%
  ggplot(aes(x=PY, y=sum, fill=Theme)) + geom_area() + 
  theme_bw() + labs(x="Year", y="Number of Publications") + 
  scale_x_continuous(breaks=seq(1990,2015, by=5), 
                     limits=c(1989, 2018), expand=c(0,0)) + 
  scale_y_continuous(limits=c(0,470), expand=c(0,0)) + 
  scale_fill_manual(values=c('darkgrey','#1b9e77','#d95f02','#7570b3','#e7298a','#66a61e'))
```
  
**Fig. S2.** Number of coastal publications over time, which are related to remote sensing, covering the entire time period (1990 - 2018).  

```{r theme, fig.width=8, fig.height=6}
# Identify papers related to biodiversity and conservation
d$`Biodiversity`[grepl(x=d$AB, pattern="biodiversity") | 
                  grepl(x=d$TI, pattern="biodiversity")] <- 1
d$`Conservation`[grepl(x=d$AB, pattern="conservation") | 
                 grepl(x=d$TI, pattern="conservation")] <- 1
d$`Climate Change`[grepl(x=d$AB, pattern="climate change") | 
                 grepl(x=d$TI, pattern="climate change")] <- 1
d$`Pollution`[grepl(x=d$AB, pattern="pollution") | 
                 grepl(x=d$TI, pattern="pollution")] <- 1
d$Sum <- rowSums(d[,c("Biodiversity", "Conservation", "Climate Change", "Pollution")], na.rm=T)
d$None[d$Sum!=0] <- NA
d$None[d$Sum==0] <- 1
d$Multiple[d$Sum>1] <- 1

p1 <- d %>% select(`Biodiversity`, `Conservation`, `Climate Change`, `Pollution`, Multiple) %>% 
  summarise_all(sum, na.rm=T) %>% tidyr::gather(var, value) %>% 
  mutate(var = factor(var, levels=c("Biodiversity", "Climate Change", "Conservation", "Pollution", "Multiple"))) %>% 
  ggplot(aes(x=var, y=value)) + 
  geom_bar(stat="identity", fill="#56B4E9") + theme_bw() + 
  geom_text(aes(label= value), hjust=-0.5) + ggtitle("a)") + 
  labs(x="", y="Number of Publications") + coord_flip() + 
  scale_y_continuous(limits=c(0,350), expand=c(0,0))

# Identify papers related to biodiversity and conservation
d$`Sea level rise`[grepl(x=d$AB, pattern="sea level rise") | 
                  grepl(x=d$TI, pattern="sea level rise")] <- 1
d$`Drought`[grepl(x=d$AB, pattern="drought") | 
                 grepl(x=d$TI, pattern="drought")] <- 1
d$`Storm intensity`[grepl(x=d$AB, pattern="storm intensity") | 
                 grepl(x=d$TI, pattern="storm intensity")] <- 1
d$`Shore line change`[grepl(x=d$AB, pattern="shore line change") | 
                 grepl(x=d$TI, pattern="shore line")] <- 1
#d$Sum <- rowSums(d[,c("Sea level rise", "Drought", "Storm intensity", "Shore line change")], na.rm=T)
#d$Multiple[d$Sum>1] <- 1

p2 <- d %>% select(`Sea level rise`, `Drought`, `Storm intensity`, `Shore line change`) %>% 
  summarise_all(sum, na.rm=T) %>% tidyr::gather(var, value) %>% 
  ggplot(aes(x=var, y=value)) + geom_bar(stat="identity", fill="#56B4E9") + theme_bw() + 
  geom_text(aes(label= value), hjust=-0.5) + ggtitle("b)") + 
  labs(x="", y="Number of Publications") + coord_flip() + 
  scale_y_continuous(limits=c(0,90), expand=c(0,0))

p1 + p2
```

**Fig. S3.** a) Number of publications by research theme (Pollution, Conservation, Climate Change, Biodiversity) and where multiple themes are covered. `r sum(d$None, na.rm=T)` publications did not cover any of the research themes considered. b) Number of publications by climate change thematic. <!-- Add lines from a) to b) see ggplot2-extension package on Github -->

```{r pub_journal, fig.width=10, fig.height=7}
d %>% select(PY, SO) %>% group_by(PY, SO) %>% summarise(sum = n()) %>% 
  tidyr::drop_na() %>% ungroup() %>% filter(sum > 5) %>% 
  mutate(PY = cut(PY, breaks=c(1989, 1995, 2000, 2005, 2010, 2015, 2020))) %>%
  ggplot(aes(x=factor(SO), y=sum, fill=PY)) + geom_bar(position="stack", stat="identity") + 
  scale_y_continuous(limits=c(0,160), breaks=seq(0,160,by=20), expand=c(0,0)) + 
  scale_fill_manual(name="Year", values=c('#7570b3','#e7298a','#66a61e','#e6ab02')) + 
  labs(x="", y="Number of Publications") + theme_bw() + coord_flip()
```

**Fig. S4.** Number of coastal publications, which are related to remote sensing, per journal and year. Only journals were more than 5 publications were found are shown.  

<!--
```{r scale, fig.width=10, fig.height=6, eval=F}
# Get scale from satellite specifications!!!

# Identify studies with information about spatial scale and add manually to dataframe
#d$TI[grep(x=d$TI, pattern="km2")]
d$scale[grepl(x=d$TI, pattern="km2")] <- c("NA")
#d$AB[grep(x=d$AB, pattern="km2")]
d$scale[grep(x=d$AB, pattern="km2")] <- c(4140, 2500, 55000, 3100, 334000, "NA", 10, 10, 8.45, 620,
                                          3400000, "NA", 25, "NA", 210, "NA")
# Duplicate number 7 and 8 are identical!!!

#Define cuts
d$scale <- cut(as.numeric(d$scale), c(0, 50, 250, 1000, 10000, 50000))

# Make plot
p1 <- d %>% select(scale) %>% tidyr::drop_na() %>%
  ggplot(aes(x=scale)) + geom_bar(fill="#56B4E9") + theme_bw() + 
  labs(x="Scale (km2)", y="Number of Publications") + 
  geom_text(stat="count", aes(label=..count..), vjust=-0.5) + 
  scale_y_continuous(limits=c(0,5), expand=c(0,0)) +
  ggplot2::annotate("text", label = "a)", x = 0.75*4/3.5, y = 5*0.95, size = 6, colour = "black")

# Resolution in meters
d$res[d$sentinel == 1] <- 10 # 5 - 60 m
d$res[d$landsat == 1] <- 30 # 30 - 60 m
d$res[d$modis == 1] <- 250 # 250 - 500 m
d$res[d$rapideye == 1] <- 6.5
d$res[d$envisat == 1] <- 300 # Ocean: 1040 m x 1020 m, Land & coast: 260m x 300m 
d$res[d$pleiades == 1] <- 0.5
d$res[d$spot == 1] <- 6
d$res[d$seawifs == 1] <- 1000 # 1.1 km LAC, 4.5 km GAC
d$res[d$aquarius == 1] <- 100000
d$res[d$grace == 1] <- 300000
d$res[d$goci == 1] <- 500
d$res[d$ikonos == 1] <- 3.2 # 1-4 m
d$res[d$aster == 1] <- 15

# What about temporal resolution?
  
#Define cuts
d$res <- cut(as.numeric(d$res), c(0, 50, 250, 1000, 10000, 50000))

p2 <- d %>% select(res) %>% tidyr::drop_na() %>%
  ggplot(aes(x=res)) + geom_bar(fill="#56B4E9") + theme_bw() + 
  labs(x="Resolution (m)", y="Number of Publications") + 
  geom_text(stat="count", aes(label=..count..), vjust=-0.5) + 
  scale_y_continuous(limits=c(0,1750), expand=c(0,0)) +
  ggplot2::annotate("text", label = "b)", x = 0.75, y = 1750*0.95, size = 6, colour = "black")
p1 + p2 + plot_layout(nrow=1)
```
**Fig. S5.** Number of studies per spatial scale (a) and spatial resolution (b).  -->
<!-- Subset small spatial resolution further, with inset graph -->

